Unsupervised training of maximum-entropy models for lexical selection in rule-based machine translation

### 2015-02-18 2015-eamt-melsa/2015-eamt-melsa.tex
some removal of comments
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58991 72bbbca6-d526-0410-a7d9-f06f51895060
more fixes
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58996 72bbbca6-d526-0410-a7d9-f06f51895060

Later, \cite{berger1996} illustrated the use of maximum-entropy
classifiers on the specific problem of lexical selection in IBM-style
word-based statistical machine translation. Other authors
\citep{melero07a} have used target-language models to
rank %@@@ carbonell06
the translations resulting from all possible combinations of lexical
selections. Nowadays, in state-of-the-art phrase-based
statistical % \todo{METIS too?}
machine translation \citep{koehn2009statistical}, lexical selection is
taken care of by a combination of the translation model and the
language model. The translation model provides probabilities of
translation between words or word sequences (often referred to as
\emph{phrases}) in the source and target language. The target-language
model provides probabilities of word sequences in the target language.
% As regards rule-based machine translation, \cite{sanchez07a}
% proposed to use information from the target language to train a
% statistical lexical-selection module but did not provide any
% experiments.  @@@ \todo[inline]{FMT: integrate \citep{sanchez07a}
% somewhere here}
\cite{marechek10} trained a maximum-entropy lexical selector for their
TectoMT rule-based system using a bilingual corpus.\todo{Say a bit
  more about TectoMT? FSM: Yes, please!}  More recently,
\cite{tyers12a} presented a method of lexical selection for rule-based
machine translation based on rules which select or remove translations
in fixed-length contexts, along with a training method for learning
the rules from a word-aligned parallel corpus.

@@

Let the probability of a word $t$ being the translation of a
word $s$ in an SL context $c$ be $p_s(t|c)$. In principle, this value
could be estimated directly from the available corpora for every
combination of $(s, t, c)$.  This would however present two questions:
(1) how should the relevant contexts be chosen? and (2) what should be
done when the translations of words are not found in the corpus?
\todo{Wording? When $(s,t,c)$ not found in the corpus?}  A
maximum-entropy model answers both of these questions. It allows the
contexts that we consider to be linguistically interesting to be
defined \emph{a priori} and then integrate these seamlessly into a
probabilistic model \citep{Manning99b}.  In answer to the second
question, a maximum-entropy model assumes nothing about what is not
present in the training data. That is, if there is no information in
the training data, then it assumes that all outcomes ---that is, all
possible translations--- are equally likely.  As previously mentioned,
the principle of maximum entropy has been applied to the problem of
lexical selection before; in particular, \cite{berger1996} cast the
problem of lexical selection in statistical MT as a classification
problem. They learn a separate maximum-entropy classifier for each SL
word form,
%\comment{[:we do too, integrate this later]} 
using SL context to
distinguish between possible translations. These classifiers are then
incorporated into the translation model of their word-based SMT
system.
%\comment{[how is our work different?]}  
In their approach, a
classifier consists of a set of binary feature functions and
corresponding weights for each feature. 
In both \cite{berger1996} and our method, features are defined in the
form $h_k^{s}(t, c)$, where $t$
is a translation, and $c$ is an SL context. One difference is that \cite{berger1996} take $s$, $t$
and $c$ to be based on word forms, whereas in our method they are based
on lemma forms.
A feature where the Spanish word \emph{pez} (`fish' as a living animal) is seen as the
translation of \emph{arrain} (`fish') in the context \emph{arrain
  handi} `big fish' would therefore be defined as:
%\comment{FSM: I have introduced the index $k$
%  because several features are defined for a SL word $s$.} 
\begin{equation}
h_k^{\mathrm{arrain}}(t, c) = \left \{ \begin{matrix}
                 ~~1 & \mbox{if } 
\left\{ \begin{array}{c}
t~ =~ \textit{pez}~ \\ \mathrm{and}~ \\ \textit{handi}~ \mathrm{follows}~ \textit{arrain} \end{array} \right. \\
                 ~~0 & \mbox{otherwise} \end{matrix} \right.
\label{eq:feat-func}
\end{equation}
This feature considers a context of zero words to the left of the
problem word and one word to the right of it.

@@

The approaches by \cite{berger1996} and by \cite{marechek10} cited
above both take advantage of a parallel corpus to collect counts of
contexts and translations in order to train maximum-entropy
models. However, parallel corpora are not available for the majority
of the world's written languages. In this section we describe an
unsupervised method to learn the models using only monolingual corpora
and the components from the RBMT system in which they are
used. \todo{FSM: Check rewording!}

The input to our method consists of a collection of samples,
$\mathcal{G} = (S, G)$, where $S = (s_1, s_2, \ldots s_{|S|})$ is a
sequence of SL words, and $G = \{g_1, g_2, \ldots g_{|G|}\}$ is a set
of possible \emph{lexical-selection paths}. A lexical-selection path
$g = (t_1, t_2, \ldots, t_{|S|})$ is a sequence of lexical-selection
choices of those SL words, where \(t_i \in
T_s(s_i)\). % and \(T_s(\cdot)\)
\todo{Check change \(T(\cdot)\to T_s(\cdot)\) for consistency with
  later notation. FSM: Check \(t_i \in T_s(s_i)\); before it was \(t_i
  = T_s(s_i)\)} %is the set of all
%translations for a source word. 
This is produced in the first stages of RBMT, just after morphological
analysis, part-of-speech tagging, and bilingual dictionary lookup, and
before any structural transfer takes place (we will call this
\emph{pre-lexsel}). In our model, it is after these first stages that
lexical selection (\emph{lexsel}) occurs. After lexical selection,
structural transfer and generation take place; a function
$\tau(g_i,S)$ represents the result of these last stages, which we
will call \emph{post-lexsel}, and returns a finished translation of a
specific lexical-selection path $g_i$ of sentence
$S$. Figure~\ref{fig:lexselschema} shows this process
schematically. %%%\todo{Can we try to bring the figure closer?}

@@

number $p(g_i|S)$ of times. We prune features (contexts)
\todo{features or contexts?} occurring less than a certain number of

@@

Apertium is a free/open-source RBMT platform, it comprises an engine,
a toolbox and data to build RBMT systems. Translation is implemented
as a pipeline consisting of the following modules: morphological
analysis, morphological disambiguation, lexical transfer, lexical
selection, structural transfer and morphological generation. The
relevant modules of the system for this method are the \emph{lexical
  transfer} module which provides for each source language lexical
form a set of possible target language lexical forms and the
\emph{structural transfer} and \emph{morphological generation} modules
which for each pair of source language lexical form and translation
produce a final translation in the target language. \todo{The wording
  of this last sentence is unclear. Identify pre-lexsel, lexsel and
  post-lexsel as before}

@@

\item[English--Spanish:] The English--Spanish pair was developed from
  a combination of the English--Catalan and Spanish--Catalan pairs
  over a period of around 3--5 months.\todo{FSM: 3, 4 or 5?} The pair
%has been used in other
%  lexical selection experiments \citep{sanchez07a}\comment{FSM: Sure?
%    It was on en-ca and in that paper no experimental results are
%    shown. This paper would be better mentioned in the introduction
%    (it mentions the use of TL info for lexical selection.},
  contains a number of entries in the bilingual dictionary with more
  than one translation. The most-frequent translation, as judged by
  the language pair developer, is marked as the \emph{default}
  translation and non-default translations were added according to
  frequency. \todo{Default is explained for some LPs only, why?}  The
  version used in this paper is SVN revision
  41387.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-en-es}}
\end{description}

@@

This section describes the measures that will be used to evaluate the
performance of the lexical selection method proposed here: a
(intrinsic) \emph{lexical selection performance} measure and an
(extrinsic) \emph{machine translation performance} measure. \todo{FSM:
I moved some stuff from here to the LER subsection.}

@@

  \begin{table}
  \begin{center}
   \begin{tabular}{|l|r|r|r|r|}
      \hline 
      %{\bf Pair}      & {\bf Lines} & {\bf SL words} & {\bf TL words} & {\bf Amb. words} & {\bf \% ambig} \\ 
      {\bf Pair}      & {\bf SL} & {\bf TL} & {\bf Amb.} & {\bf \% amb.} \\ 
      \hline 
       \texttt{br-fr} & 13,854  & 13,878  & 1,163 & 8.39   \\
      \hline 
       \texttt{mk-en} & 13,441 & 14,228  & 3,872 & 28.80 \\
      \hline 
       \texttt{eu-es} & 7,967 & 11,476  & 1,360  & 17.07 \\
      \hline 
       \texttt{en-es} & 19,882 & 20,944  & 1,469  & 7.38 \\
      \hline 
   \end{tabular} 
  \end{center}
   \caption{Statistics about the test corpora. The columns \textbf{SL} and \textbf{TL} give the number
      of tokens in the source and target languages respectively. The columns \textbf{amb.\ words} and \textbf{\%~ambig} gives 
      the number of word with more than one translation and the percentage of SL words which have 
      more than one translation respectively.}
    \label{table:test-corp}
  \end{table}

  After this process, we selected 1,000 sentence pairs at random for
  testing (\texttt{test}), 1,000 for development
  (\texttt{dev})\footnote{The development corpus was used for checking
    the value for frequency pruning of features.} and left the
  remainder for training.
%This is a smaller
%number of sentences than used in typical evaluations (such as the WMT
%series, \cite{callisonburch2012}), but was motivated by the fact that
%the smallest corpus (for Breton--French) after sentence discarding was
%only little over 4,000 sentences long. 
  Table~\ref{table:input-corp} gives statistics about the size of the
  input corpora, and how many sentences were left after processing for
  testing, training and development. \todo{Why do ambiguity numbers
    change so much from the paper?} Table~\ref{table:test-corp} gives
  information about the test corpora. \todo{FSM: One may wonder which
    would be the results when the TL side of the training corpus is
    used to train the models in the usual way.}

@@

As we are working with binary features, we use the implementation of
generalised iterative scaling available in the {\sc
  yasmet}\footnote{\url{http://www-i6.informatik.rwth-aachen.de/web/Software/YASMET.html};
  the compilable version we used is available as part of the Apertium
  \texttt{lex-tools} package,
  \url{http://downloads.sourceforge.net/project/apertium/apertium-lex-tools/apertium-lex-tools-0.1.0.tar.gz}.}
to calculate the feature weights.  After learning the feature sets and
weights, we compute the evaluation measures described in
section~\ref{ss:measures}. There is an option to remove features
(contexts)\todo{FSM: Features are contexts, aren't they?} which occur
less than a certain number of times in the training corpus. This is
referred to as the feature pruning frequency threshold --- features
occuring less than the threshold are discarded. The value was set
experimentally. Values of between two and seven were tested, and the
ones which provided the best improvement on the development corpus
were selected; they happen to come close to the rule-of-thumb value of
five that \citet[p.~596]{Manning99b} found to be effective.
Table~\ref{table:maxent-features} shows the number of features that
have eventually been used for each language pair.

@@

%The authors thank Filip Petkovski for his help in running some of the
%experiments. They also acknowledge partial support from the 
%Spanish Government through projects 
%TIN2009-14009-C02-01 and TIN2012-32615, the European Commission through project
%Abu-MaTran (FP7-PEOPLE, 324414) (and the NILS mobility grant (Abel Predoc
%Research Grant) coordinated by the Universidad Complutense de Madrid.
%
\bibliographystyle{apalike}
\small{
\bibliography{2015-eamt-melsa}
}

>>>>

Later, \cite{berger1996} illustrated the use of maximum-entropy
classifiers on the specific problem of lexical selection in IBM-style
word-based statistical machine translation. Other authors
\citep{melero07a} have used target-language models to
rank %@@@ carbonell06
the translations resulting from all possible combinations of lexical
selections. Nowadays, in state-of-the-art phrase-based
statistical % \todo{METIS too?}
machine translation \citep{koehn2009statistical}, lexical selection is
taken care of by a combination of the translation model and the
language model. The translation model provides probabilities of
translation between words or word sequences (often referred to as
\emph{phrases}) in the source and target language. The target-language
model provides probabilities of word sequences in the target language.
% As regards rule-based machine translation, \cite{sanchez07a}
% proposed to use information from the target language to train a
% statistical lexical-selection module but did not provide any
% experiments.  @@@ \todo[inline]{FMT: integrate \citep{sanchez07a}
% somewhere here}
\cite{marechek10} trained a maximum-entropy lexical selector for their
dependency-grammar-based transfer system TectoMT using a bilingual corpus.
More recently, \cite{tyers12a} presented a method of lexical selection for rule-based
machine translation based on rules which select or remove translations
in fixed-length contexts, along with a training method for learning
the rules from a word-aligned parallel corpus.

@@

Let the probability of a word $t$ being the translation of a
word $s$ in an SL context $c$ be $p_s(t|c)$. In principle, this value
could be estimated directly from the available corpora for every
combination of $(s, t, c)$.  This would however present two questions:
(1) how should the relevant contexts be chosen? and (2) what should be
done when $(s, t, c)$ is not found in the corpus?
A maximum-entropy model answers both of these questions. It allows the
contexts that we consider to be linguistically interesting to be
defined \emph{a priori} and then integrate these seamlessly into a
probabilistic model \citep{Manning99b}.  In answer to the second
question, a maximum-entropy model assumes nothing about what is not
present in the training data. That is, if there is no information in
the training data, then it assumes that all outcomes ---that is, all
possible translations--- are equally likely.  As previously mentioned,
the principle of maximum entropy has been applied to the problem of
lexical selection before; in particular, \cite{berger1996} cast the
problem of lexical selection in statistical MT as a classification
problem. They learn a separate maximum-entropy classifier for each SL
word form,
%\comment{[:we do too, integrate this later]} 
using SL context to
distinguish between possible translations. These classifiers are then
incorporated into the translation model of their word-based SMT
system.
%\comment{[how is our work different?]}  
In their approach, a
classifier consists of a set of binary feature functions and
corresponding weights for each feature. 
In both \cite{berger1996} and our method, features are defined in the
form $h_k^{s}(t, c)$, where $t$
is a translation, and $c$ is an SL context. One difference is that \cite{berger1996} take $s$, $t$
and $c$ to be based on word forms, whereas in our method they are based
on lemma forms.
A feature where the Spanish word \emph{pez} (`fish' as a living animal) is seen as the
translation of \emph{arrain} (`fish') in the context \emph{arrain
  handi} `big fish' would therefore be defined as:
%\comment{FSM: I have introduced the index $k$
%  because several features are defined for a SL word $s$.} 
\begin{equation}
h_k^{\mathrm{arrain}}(t, c) = \left \{ \begin{matrix}
                 ~~1 & \mbox{if } 
\left\{ \begin{array}{c}
t~ =~ \textit{pez}~ \\ \mathrm{and}~ \\ \textit{handi}~ \mathrm{follows}~ \textit{arrain} \end{array} \right. \\
                 ~~0 & \mbox{otherwise} \end{matrix} \right.
\label{eq:feat-func}
\end{equation}
This feature considers a context of zero words to the left of the
problem word and one word to the right of it.

@@

The approaches by \cite{berger1996} and by \cite{marechek10} cited
above both take advantage of a parallel corpus to collect counts of
contexts and translations in order to train maximum-entropy
models. However, parallel corpora are not available for the majority
of the world's written languages. In this section we describe an
unsupervised method to learn the models using only monolingual corpora
and the components from the RBMT system in which they are
used.

The input to our method consists of a collection of samples,
$\mathcal{G} = (S, G)$, where $S = (s_1, s_2, \ldots s_{|S|})$ is a
sequence of SL words, and $G = \{g_1, g_2, \ldots g_{|G|}\}$ is a set
of possible \emph{lexical-selection paths}. A lexical-selection path
$g = (t_1, t_2, \ldots, t_{|S|})$ is a sequence of lexical-selection
choices of those SL words, where \(t_i \in
T_s(s_i)\). % and \(T_s(\cdot)\)
%is the set of all
%translations for a source word. 
This is produced in the first stages of RBMT, just after morphological
analysis, part-of-speech tagging, and bilingual dictionary lookup, and
before any structural transfer takes place (we will call this
\emph{pre-lexsel}). In our model, it is after these first stages that
lexical selection (\emph{lexsel}) occurs. After lexical selection,
structural transfer and generation take place; a function
$\tau(g_i,S)$ represents the result of these last stages, which we
will call \emph{post-lexsel}, and returns a finished translation of a
specific lexical-selection path $g_i$ of sentence
$S$. Figure~\ref{fig:lexselschema} shows this process
schematically. %%%\todo{Can we try to bring the figure closer?}

@@

number $p(g_i|S)$ of times. We prune $(s, t, c)$
occurring less than a certain number of

@@

Apertium is a free/open-source RBMT platform, it comprises an engine,
a toolbox and data to build RBMT systems. Translation is implemented
as a pipeline consisting of the following modules: morphological
analysis, morphological disambiguation, lexical transfer, lexical
selection, structural transfer and morphological generation. 
%The relevant modules of the system for this method are the \emph{lexical
%  transfer} module which provides for each source language lexical
%form a set of possible target language lexical forms and the
%\emph{structural transfer} and \emph{morphological generation} modules
%which for each pair of source language lexical form and translation
%produce a final translation in the target language. \todo{The wording
%  of this last sentence is unclear. Identify pre-lexsel, lexsel and
%  post-lexsel as before}

@@

\item[English--Spanish:] The English--Spanish pair was developed from
  a combination of the English--Catalan and Spanish--Catalan pairs
  over a period of around four months. The pair
%has been used in other
%  lexical selection experiments \citep{sanchez07a}\comment{FSM: Sure?
%    It was on en-ca and in that paper no experimental results are
%    shown. This paper would be better mentioned in the introduction
%    (it mentions the use of TL info for lexical selection.},
  contains a number of entries in the bilingual dictionary with more
  than one translation. %The most-frequent translation, as judged by
%  the language pair developer, is marked as the \emph{default}
%  translation and non-default translations were added according to
%  frequency. \todo{Default is explained for some LPs only, why?}  
  The
  version used in this paper is SVN revision
  41387.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-en-es}}
\end{description}

@@

This section describes the measures that will be used to evaluate the
performance of the lexical selection method proposed here: a
(intrinsic) \emph{lexical selection performance} measure and an
(extrinsic) \emph{machine translation performance} measure.

@@

  %% FMT: INCLUDE AVERAGE AMBIGUITY OVER AMBIGUOUS WORDS
  \begin{table}
  \begin{center}
   \begin{tabular}{|l|r|r|r|r|}
      \hline 
      %{\bf Pair}      & {\bf Lines} & {\bf SL words} & {\bf TL words} & {\bf Amb. words} & {\bf \% ambig} \\ 
      {\bf Pair}      & {\bf SL} & {\bf TL} & {\bf Amb.} & {\bf \% amb.} \\ 
      \hline 
       \texttt{br-fr} & 13,854  & 13,878  & 1,163 & 8.39   \\
      \hline 
       \texttt{mk-en} & 13,441 & 14,228  & 3,872 & 28.80 \\
      \hline 
       \texttt{eu-es} & 7,967 & 11,476  & 1,360  & 17.07 \\
      \hline 
       \texttt{en-es} & 19,882 & 20,944  & 1,469  & 7.38 \\
      \hline 
   \end{tabular} 
  \end{center}
   \caption{Statistics about the test corpora. The columns \textbf{SL} and \textbf{TL} give the number
      of tokens in the source and target languages respectively. The columns \textbf{amb.\ words} and \textbf{\%~ambig} gives 
      the number of word with more than one translation and the percentage of SL words which have 
      more than one translation respectively.}
    \label{table:test-corp}
  \end{table}

  After this process, we selected 1,000 sentence pairs at random for
  testing (\texttt{test}), 1,000 for development
  (\texttt{dev})\footnote{The development corpus was used for checking
    the value for frequency pruning of features.} and left the
  remainder for training.
%This is a smaller
%number of sentences than used in typical evaluations (such as the WMT
%series, \cite{callisonburch2012}), but was motivated by the fact that
%the smallest corpus (for Breton--French) after sentence discarding was
%only little over 4,000 sentences long. 
  Table~\ref{table:input-corp} gives statistics about the size of the
  input corpora, and how many sentences were left after processing for
  testing, training and development. \todo{Why do ambiguity numbers
    change so much from the paper?} Table~\ref{table:test-corp} gives
  information about the test corpora. 
%! \todo{FSM: One may wonder which
%!    would be the results when the TL side of the training corpus is
%!    used to train the models in the usual way.}

@@

As we are working with binary features, we use the implementation of
generalised iterative scaling available in the {\sc
  yasmet}\footnote{\url{http://www-i6.informatik.rwth-aachen.de/web/Software/YASMET.html};
  the compilable version we used is available as part of the Apertium
  \texttt{lex-tools} package,
  \url{http://downloads.sourceforge.net/project/apertium/apertium-lex-tools/apertium-lex-tools-0.1.0.tar.gz}.}
to calculate the feature weights.  After learning the feature sets and
weights, we compute the evaluation measures described in
section~\ref{ss:measures}. There is an option to remove events 
$(s, t, c)$ which occur
less than a certain number of times in the training corpus. This is
referred to as the feature pruning frequency threshold --- features
occuring less than the threshold are discarded. The value was set
experimentally. Values of between two and seven were tested, and the
ones which provided the best improvement on the development corpus
were selected; they happen to come close to the rule-of-thumb value of
five that \citet[p.~596]{Manning99b} found to be effective.
Table~\ref{table:maxent-features} shows the number of features that
have eventually been used for each language pair.

@@

%The authors thank Filip Petkovski for his help in running some of the
%experiments. They also acknowledge partial support from the 
%Spanish Government through projects 
%TIN2009-14009-C02-01 and TIN2012-32615, the European Commission through project
%Abu-MaTran (FP7-PEOPLE, 324414) (and the NILS mobility grant (Abel Predoc
%Research Grant) coordinated by the Universidad Complutense de Madrid.

\bibliographystyle{apalike}
\small{
\bibliography{2015-eamt-melsa}
}
